Automating Free Logic in HOL, with an Experimental Application in Category Theory

### 2018-10-10 FreeLogic/2017-Journal/REVISED/article.tex
added

  A shallow semantical embedding of free logic in classical
  higher-order logic is presented, which enables the off-the-shelf
  application of higher-order interactive and automated theorem
  provers for the formalisation and
  verification of free logic theories.  
  Subsequently, this approach is exemplarily employed in a selected domain of
  mathematics: starting from a generalization of the standard axioms
  for a monoid we present a stepwise development of various, mutually
  equivalent foundational axiom systems for category theory. As a
  side-effect of this work some (minor) issue in a prominent category
  theory textbook has been revealed.

  The purpose of this article is not to claim any novel results in
  category theory, but to demonstrate an elegant way to ``implement''
  and utilize interactive and automated reasoning in free logic, and
  to present respective experiments.

@@

In the first part of this article (\S2 and \S3) we show how free logic
can be elegantly ``implemented'' in any theorem proving system for
classical higher-order logic (HOL) \cite{B5}. The proposed solution
employs a semantic embedding of free in HOL. We present, as an
example, one implementation of this idea in the mathematical proof
assistant Isabelle/HOL \cite{NPW02}. Various state-of-the-art
first-order and higher-order automated theorem provers and model
finders are integrated (modulo suitable logic translations) with
Isabelle via the Sledgehammer tool \cite{Sledgehammer}, so that our
solution can be utilized, via Isabelle as foreground system, with a
whole range of other background reasoners. As a result we obtain an
elegant and powerful implementation of an interactive and automated
theorem proving (and model finding) system for free logic.

@@

\section{Summary and Further Work}
We have developed a new reasoning framework for free logic, and we
have experimentally applied it for some first experiments in category
theory.  We have demonstrated how modern proof assistants and theorem
provers for classical higher-order logic may well support the
reasoning in free logic. More concretely, we have applied our new free
logic reasoning framework for the systematic exploration of axiom
systems for category theory. Without tools, support of such experiments
would be extremely tedious and error prone.
In the course of our experiments, automated theorem provers 
have revealed some (minor) issue in the textbook of Freyd and Scedrov \cite{FreydScedrov90},
which we were able to correct. The correction essentially corresponds
to the axiom system by Scott proposed earlier \cite{Scott67}.
All our findings were achieved directly by or in
close interaction with automated reasoning tools.

>>>>

  A shallow semantical embedding of free logic in classical
  higher-order logic is presented, which enables the off-the-shelf
  application of higher-order interactive and automated theorem
  provers for the formalisation and
  verification of free logic theories.  
  Subsequently, this approach is applied to a selected domain of
  mathematics:
 starting from a generalization of the standard axioms
  for a monoid we present a stepwise development of various, mutually
  equivalent foundational axiom systems for category theory. As a
  side-effect of this work some (minor) issue in a prominent category
  theory textbook has been revealed.

  The purpose of this article is not to claim any novel results in
  category theory, but to demonstrate an elegant way to ``implement''
  and utilize interactive and automated reasoning in free logic, and
  to present illustrative experiments.

@@

In the first part of this article (\S2 and \S3) we show how free logic
can be elegantly ``implemented'' in any theorem proving system for
classical higher-order logic (HOL) \cite{B5}. The proposed solution
employs a semantic embedding of free logic in HOL. We present, as an
example, one implementation of this idea in the mathematical proof
assistant Isabelle/HOL \cite{NPW02}. Various state-of-the-art
first-order and higher-order automated theorem provers and model
finders are integrated (modulo suitable logic translations) with
Isabelle via the Sledgehammer tool \cite{Sledgehammer}, so that our
solution can be utilized, via Isabelle as foreground system, with a
whole range of other background reasoners. As a result we obtain an
elegant and powerful implementation of an interactive and automated
theorem proving (and model finding) system for free logic.

@@

\section{Summary and Further Work}
We have developed a new reasoning framework for free logic, and we
have experimentally applied it for some first experiments in category
theory.  We have demonstrated how modern proof assistants and theorem
provers for classical higher-order logic may well support the
reasoning in free logic. More concretely, we have applied our new free
logic reasoning framework for the systematic exploration of axiom
systems for category theory. Without tools, support of such experiments
would be extremely tedious and error prone.
In the course of our experiments, automated theorem provers 
have revealed some (minor) issue in the textbook of Freyd and Scedrov \cite{FreydScedrov90},
which we were able to correct. The correction essentially corresponds
to the axiom system by Scott proposed earlier \cite{Scott67}.
All our findings were achieved directly by or in
close interaction with automated reasoning tools. 
Perhaps the lesson to be learned here is that, when working with partial functions,
it is natural --- out of caution  --- to assume too much, and the automated reasoning
tools, as we have shown here, can help find in what ways the axioms might be reduced or
simplified.
